---
title: "A First Look at mtcars"
author: "Ruud Kooiman"
date: today
format:
  html:
    toc: true
    code-fold: show
    theme: cosmo
execute:
  warning: false
---

## Introduction

This is a short test document written in Quarto using the R (knitr) engine. It
uses `mtcars`, the dataset of 32 cars from a 1974 issue of *Motor Trend*, to
show the three things a Quarto document usually needs to do: run code, draw a
figure, and print a table.

The question it looks at is a simple one: how does the weight of a car relate to
its fuel economy?

## The data

```{r}
#| label: load-data

data(mtcars)
dim(mtcars)
head(mtcars[, c("mpg", "cyl", "hp", "wt")])
```

The dataset has `r nrow(mtcars)` cars and `r ncol(mtcars)` variables. Below we
use four of them: `mpg` (miles per gallon), `cyl` (number of cylinders), `hp`
(gross horsepower) and `wt` (weight in thousands of pounds).

## A plot

@fig-weight-mpg plots fuel economy against weight, with the number of cylinders
shown by colour and a least-squares line through all points.

```{r}
#| label: fig-weight-mpg
#| fig-cap: "Fuel economy falls as weight rises; heavier cars also tend to have more cylinders."
#| fig-alt: "Scatter plot of miles per gallon against car weight for 32 cars, coloured by cylinder count, with a downward-sloping regression line."
#| fig-width: 7
#| fig-height: 4.5

cyl_levels <- sort(unique(mtcars$cyl))
palette_cyl <- c("#2c7fb8", "#d95f02", "#1b9e77")

plot(
  mtcars$wt, mtcars$mpg,
  col  = palette_cyl[match(mtcars$cyl, cyl_levels)],
  pch  = 19,
  cex  = 1.3,
  xlab = "Weight (1000 lbs)",
  ylab = "Miles per gallon",
  main = "Weight versus fuel economy"
)

abline(lm(mpg ~ wt, data = mtcars), col = "grey40", lwd = 2, lty = 2)

legend(
  "topright",
  legend = paste(cyl_levels, "cylinders"),
  col    = palette_cyl,
  pch    = 19,
  bty    = "n"
)
```

## A summary table

@tbl-by-cyl summarises the same three variables per cylinder group.

```{r}
#| label: tbl-by-cyl
#| tbl-cap: "Mean values per cylinder group."

summary_tbl <- aggregate(
  cbind(mpg, hp, wt) ~ cyl,
  data = mtcars,
  FUN  = mean
)
summary_tbl$n <- as.vector(table(mtcars$cyl))

knitr::kable(
  summary_tbl[, c("cyl", "n", "mpg", "hp", "wt")],
  col.names = c("Cylinders", "Cars", "Mean mpg", "Mean hp", "Mean weight"),
  digits    = 1
)
```

## Conclusion

Both the figure and the table point the same way: more cylinders go together
with more weight and more horsepower, and with markedly worse fuel economy —
from about 27 mpg for the four-cylinder cars down to about 15 mpg for the
eight-cylinder ones.
